Monte Carlo Synthetic Acceleration Solutions for the Simplified $P_N$ Approximation

### 2014-12-04 doc/ttsp2013/paper/ttsp_2013.tex
cleaning up abstract and introduction in ttsp paper

We present Monte Carlo Synthetic Acceleration (MCSA) as a linear
solver technique for the simplified $P_N$ ($SP_N$) discretization of
the Boltzmann neutron transport equation. Using the fully-formed
linear operators for the transport problem, we explore solutions to
the $SP_N$ equations with MCSA using criticality calculations as the
driving problems. Due to neutron scattering in the light water
moderator, several difficulties arise when applying MCSA to these
problems that were not observed in previous work where the method was
applied to more opaque thermal radiaton transport systems. Of most
significance, it was discovered for these problems that simply
reducing the spectral radius of the Monte Carlo problem below unity is
not enough to efficiently achieve convergence as reported in previous
work. In order to effectively solve the $SP_N$ equations for the model
problem, an algebraic preconditioning strategy is developed for MCSA
where we utilize sparse approximate inverse representations of the
operator as preconditioners to achieve convergence. Using the
preconditioners, MCSA performance is compared with production
implementations of preconditioned Krylov methods. Our studies indicate
that the iterative performance of MCSA when used to solve the $SP_N$
equations for light water reactor problems is competitive with the
preconditioned Krylov methods when operator spectrum is properly
conditioned. To solve large problems, we have parallelized the method
by constructing a variant of the multiple-set overlapping-domain
(MSOD) algorithm for Monte Carlo calculations. We discuss the method
and then present results indicating its scalability.

@@

Predictive modeling and simulation capability requires the combination
of high fidelity models, high performance computing hardware that can
handle the intense computational loads required by these models, and
modern algorithms for solving these problems that leverage this high
performance hardware. For nuclear reactor analysis, this predictive
capability can enable tighter design tolerances for improved thermal
performance and efficiency, higher fuel utilization and burnup, and
high confidence in accident scenario models. Modern high fidelity
deterministic methods for large scale problems where the solution over
the entire core of a reactor is desired are often variants on the
discrete ordinates ($S_N$) method \cite{evans_denovo:_2010} or the
method of characteristics \cite{askew_1972}. For fission reactor
neutronics simulations, the discrete ordinates method requires
potentially trillions of unknown angular flux moments to be computed
to achieve good accuracy for the responses of interest
\cite{slaybaugh_acceleration_2011} which in many cases can be
intractable without sufficient computational hardware. On the other
side of the spectrum, low-fidelity diffusion solutions are often
leveraged as an alternative to compute a solution over the entire
domain a reactor core. In the 1960's, Gelbard developed an ad-hoc,
multidimensional extension of the simple single dimension planar $P_N$
equations that created a system of coupled, diffusion-like equations
known as the simplified $P_N$ ($SP_N$) equations \cite{gelbard_1960}
which incorporates additional angular information into a set of
diffusion-like equations. Further work in both aysmptotic analysis and
algorithmic development proceeded in the following decades to serve as
a theoretical basis for the method
\cite{morel_1996,brantley_simplified_2000}. Recent work has expanded
the discretization to a fully algebraic form ammenable to solution
with modern linear algebra techniques including Krylov solvers and
advanced precondtioners including multigrid methods and incomplete
factorizations \cite{hamilton_2014}. Using these new techniques, the
reduction in numerical complexity compared to current deterministic
solution methods using the $SP_N$ approximation could mean significant
savings in both compute time and memory required and permit larger
simulations to be performed on the available computing platforms.

Looking towards future computing platforms, as machines begin to
operate at hundreds of petaflops peak performance and beyond, trends
toward reduced energy consumption will require incredibly high levels
of concurrency to achieve the desired computation rates
\cite{kogge_using_2011}. The end result of these hardware changes is
that the larger number of low-powered processors will be prone to both
soft failures such as bit errors in floating point operations and hard
failures where the data owned by that processor cannot be recovered
\cite{u.s._department_of_energy_resilient_2012}. Because these
failures are predicted to be common, resilient solver technologies are
are a possible means of overcoming these events at the application
level. With solutions to deterministic transport problems based on
Monte Carlo techniques, such issues are potentially alleviated by
statistical arguments. In the case of soft failures, isolated floating
point errors in the Monte Carlo simulation are absorbed within tally
statistics while a hard failure is treated as a high variance event
where some portion of the Monte Carlo histories lost.

Considering these physics and hardware issues, the goal of this work
is to research and develop Monte Carlo Synthetic Acceleration methods
for neutron transport problems. These methods are based on using the
Monte Carlo method to directly accelerate the convergence of
traditional fixed-point iterative schemes for these problems. We aim
to address both their benefits and their shortcomings in the context
of the physics of interest and desire to identify areas in which
improvements can be made. In previous work \cite{evans_monte_2014}, we
applied the Monte Carlo synthetic acceleration (MCSA) method to the
thermal radiation diffusion equation. For several transient problems
we observed that MCSA performed competetively with preconditioned
Krylov methods. In a continuation of these studies, we apply the
method to light water reactor problems discretized by the $SP_N$
equations. This system has no time dependence, a large scattering
component, and assymetry in the linear operator, creating a set of
problems complementary to the transient, opaque, and symmetric systems
of our previous analyis. In addition, we apply a new fixed point
iteration in the MCSA sequence as well as a new estimator in order to
study their effects on convergence.

The paper is organized as follows. First, we briefly introduce the
$SP_N$ equations for eigenvalue problems along with the discretization
used for this work and the model problem. We then describe the
application of the MCSA algorithm to the solution of the
equations. Next, we demonstrate that the basic Jacobi preconditioning
applied in our previous work is unsatisfactory to achieve convergence
for neutron transport in light water reactors and subsquently develop
a new preconditioning scheme. We then apply this preconditioning
scheme to a set of criticality problems and compare the
performance of the preconditioned MCSA method to preconditioned Krylov
methods. We then parallelize the MCSA algorithm leveraging the MSOD
scheme to enable very large problems with a demonstration of
scalability using the test problems. We then finish with conclusions
and indication of future work.

%%---------------------------------------------------------------------------%%
\section{Simplified $P_N$ Approximation}
For criticality problems, the space, angle, and energy discretizations
for the $SP_N$ equations can be applied to the general eigenvalue form
of the transport equation given by
Eq~(\ref{eq:eigenvalue_transport}):
\begin{multline}
  -\nabla \cdot \Bigg[\frac{n}{2n+1}\mathbf{\Sigma_{n-1}}^{-1} \nabla
    \Big(\frac{n-1}{2n-1} \mathbf{\Phi_{n-2}} +
    \frac{n}{2n-1}\mathbf{\Phi_n} \Big) \\+
    \frac{n+1}{2n+1}\mathbf{\Sigma_{n+1}}^{-1} \nabla
    \Big(\frac{n+1}{2n+3}\mathbf{\Phi_n} +
    \frac{n+2}{2n+3}\mathbf{\Phi_{n+2}}\Big) \Bigg] \\+
  \mathbf{\Sigma_n} \mathbf{\Phi_n} = \frac{1}{k} \mathbf{F}
  \mathbf{\Phi_n} \delta_{n0} \ \ \ \ \ \ \ \ \ n = 0,2,4,\cdots,N\:.
  \label{eq:multigroup_spn_eigenvalue}
\end{multline}
with $\mathbf{\Phi_n}$ the vector of multigroup flux moments and the
fission matrix, $\mathbf{F}$. We apply the following change of
variables to yield the set of multigroup pseudo-moments
$\mathbb{U}_n$:
\begin{subequations}
  \begin{gather}
    \mathbb{U}_1 = \mathbf{\Phi}_0 + 2\mathbf{\Phi}_2 \\
    \mathbb{U}_2 = 3\mathbf{\Phi}_2 + 4\mathbf{\Phi}_4 \\
    \mathbb{U}_3 = 5\mathbf{\Phi}_4 + 6\mathbf{\Phi}_6 \\
    \mathbb{U}_4 = 7\mathbf{\Phi}_6 \:.
  \end{gather}
  \label{eq:spn7_subs}
\end{subequations}
where now the application of the fission matrix to the moment vectors
will be expanded into a set of block matrices in identical fashion to
the scattering matrices:
\begin{equation}
  -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
  \mathbb{A}_{nm} \mathbb{U}_m = \frac{1}{k} \sum_{m=1}^4
  \mathbb{F}_{nm} \mathbb{U}_m\ \ \ \ \ \ \ n = 1,2,3,4\:.
  \label{eq:spn_fission_matrix}
\end{equation}
To complete the spatial discretization, the gradient operators are
resolved by a finite volume scheme as in \cite{hamilton_2014} such
that the flux and current are continuous at cell boundaries.. It
should be noted here that with respect to a general MCSA scheme, the
introduction of fission in the system will only affect the source
vector in the linear system for the eigenvalue schemes used in this
work. The linear operator, and therefore overall MCSA performance will
be dictated by streaming and scattering as defined on the left-hand
side. We refer the reader to \cite{hamilton_2014} for an extensive
presentation of the $SP_N$ discretization used in this work.

%%---------------------------------------------------------------------------%%
\section{Monte Carlo Synthetic Acceleration}
\label{sec:mcsa}
In this section we briefly review the Monte Carlo method for linear
systems and Monte Carlo Synthetic Acceleration. Given a linear
problem:
\begin{equation}
  \ve{A} \ve{x} = \ve{b}\:,
  \label{eq:linear_problem}
\end{equation}
where $\ve{A} \in \mathbb{R}^{N \times N}$ is a matrix operator such
that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
\mathbb{R}^N$ is the solution vector, and $\ve{b} \in
\mathbb{R}^N$. To form the Monte Carlo method we define the linear
system adjoint to Eq~(\ref{eq:linear_problem}):
\begin{equation}
  \ve{A}^T \ve{y} = \ve{d}\:,
  \label{eq:adjoint_linear_problem}
\end{equation}
where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
vectors respectively and $\ve{A}^T$ is the adjoint operator for
$\ve{A} \in \mathbb{R}^{N \times N}$ which we hten split to get:
\begin{equation}
  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:,
  \label{eq:adjoint_split_system}
\end{equation}
where $\ve{H}^T$ is the adjoint iteration matrix. The spectral radius
of $\ve{H}^T$ must remain less than 1 for convergence. By defining the
following adjoint inner product equivalence \cite{spanier_monte_1969}:
\begin{equation}
  \langle \ve{A}^T \ve{y}, \ve{x} \rangle = \langle \ve{y}, \ve{A}
  \ve{x} \rangle\:.
  \label{eq:adjoint_operator_product}
\end{equation}
it follows that:
\begin{equation}
  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
  \label{eq:adjoint_vector_relation}
\end{equation}
We can acquire the adjoint solution by forming the Neumann series by
writing Eq~(\ref{eq:adjoint_split_system}) as:
\begin{equation}
  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
  \label{eq:adjoint_split_system_2}
\end{equation}
which in turn yields the Neumann series using the adjoint operator:
\begin{equation}
  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
  \label{eq:adjoint_neumann_series}
\end{equation}
We expand this summation to again yield a series of transitions that
can be approximated by a Monte Carlo random walk sequence by forming
the Neumann series in reverse order:
\begin{equation}
  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
  \label{eq:adjoint_neumann_solution}
\end{equation}
We can readily build an estimator for the adjoint solution from this
series expansion, but we instead desire the solution to
Eq~(\ref{eq:linear_problem}). Here we have 2 unknowns, $\ve{y}$ and
$\ve{d}$, and therefore we require two constraints to close the
system. We use Eq~(\ref{eq:adjoint_vector_relation}) as the first
constraint and as a second constraint we select:
\begin{equation}
  \ve{d} = \boldsymbol{\delta}_j\:,
  \label{eq:adjoint_second_constraint}
\end{equation}
where $\boldsymbol{\delta}_j$ is one of a set of vectors in which the
$j^{th}$ component is the Kronecker delta function $\delta_{i,j}$. If
we apply Eq~(\ref{eq:adjoint_second_constraint}) to our first
constraint Eq~(\ref{eq:adjoint_vector_relation}), we get the following
convenient outcome:
\begin{equation}
  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
  \boldsymbol{\delta}_j \rangle = x_j \:,
  \label{eq:inner_product_constraint}
\end{equation}
meaning that if we compute the inner product of the original source
and the adjoint solution using a delta function source, we recover one
component of the original solution.

In terms of radiation transport, this adjoint method is equivalent to
a traditional forward method where the initial state $i_0$ of the
random walk is determined by sampling the source vector $\ve{b}$ with
probabilities:
\begin{equation}
  P_{(i_0=i)}(\nu) = \frac{|b_i|}{||\ve{b}||_1}\:,
  \label{eq:adjoint_source_probability}
\end{equation}
with a random walk starting weight of:
\begin{equation}
  W_0 = ||\ve{b}||_1 \frac{b_i}{|b_i|}\:,
  \label{eq:adjoint_starting_weight}
\end{equation}
which gives the additional useful relation:
\begin{equation}
  b_{i_0} = W_0 P_{(i_0=i)}\:.
  \label{eq:adjoint_source_definition}
\end{equation}
We build weights and probabilities for the random walk using the
\textit{adjoint Neumann-Ulam decomposition} of $\ve{H}$:
\begin{equation}
  \ve{H}^{T} = \ve{P} \circ \ve{W}\:,
  \label{eq:adjoint_neumann_ulam}
\end{equation}
where we are forming the decomposition with respect to the transpose
of $\ve{H}$. $\ve{P}$ is defined element-wise as:
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,
  \label{eq:adjoint_probability}
\end{equation}
such that we expect to select a new state, $j$, from the current state
in the random walk, $i$, by sampling column-wise (or row-wise if an
adjoint probability matrix is formed). Per
Eq~(\ref{eq:adjoint_neumann_ulam}), the transition weight is then
defined as:
\begin{equation}
  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
  \label{eq:adjoint_weight}
\end{equation}
Using the decomposition we can then define an expectation value for
the adjoint method. The contribution to the solution in state $j$ from
a particular random walk permutation of $k$ events is given by
\textit{collision estimator}:
\begin{equation}
  X_{j}(\nu) = \sum_{m=0}^k W_{m} \delta_{i_m,j}\:,
  \label{eq:adjoint_permutation_contribution}
\end{equation}
where the Kronecker delta indicates that the tally contributes only in
the current state, $i_m$, of the random walk. The expectation value
using all permutations is then:
\begin{equation}
  E\{X_j\} = \sum_{\nu} P_{\nu} X_{j}(\nu)\:,
  \label{eq:adjoint_expectation_value}
\end{equation}
which, if expanded utilizing Eq~(\ref{eq:adjoint_source_definition})
to insert the source term, directly recovers the exact solution:
\begin{equation}
  \begin{split}
    E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N} b_{i_0} p_{i_0,i_1}p_{i_1,i_2}\ldots
    p_{i_{k-1},i_k} w_{i_0,i_1}w_{i_1,i_2}\ldots
    w_{i_{k-1},i_k} \delta_{i_k,j} \\ &= x_{j}\:,
  \end{split}
  \label{eq:adjoint_expectation_expansion}
\end{equation}
therefore, also providing an unbiased Monte Carlo estimate of the
solution. It should be noted here that
Eq~(\ref{eq:adjoint_expectation_expansion}) only computes a single
component of our desired solution vector when really what we desire is
the entire solution vector. In an adjoint Monte Carlo simulation using
this estimator, the $w_{ij}$ elements that are added into the tally
for each state are only selected if/when the random walk currently
resides in that state. Much like a mesh tally in a particle transport
simulation, we have $N$ simultaneous tallies for $\ve{A} \in
\mathbb{R}^{N \times N}$ that will yield the entire solution vector.

>>>>

We present Monte Carlo Synthetic Acceleration (MCSA) as a linear
solver technique for the simplified $P_N$ ($SP_N$) discretization of
the Boltzmann neutron transport equation. Using the fully-formed
linear operators for the transport problem, we explore solutions to
the $SP_N$ equations with MCSA using light water criticality
calculations as the driving problems. Due to neutron scattering in the
light water moderator, several difficulties arise when applying MCSA
to these problems that were not observed in previous work where the
method was applied to more opaque thermal radiaton transport
systems. In addition to numerical difficulties, the level of fidelity
needed for the accurate solution of these problems creates enough
degrees of freedom to warrant domain decomposition on hundreds to
thousands of cores in order to obtain both a time-efficient solution
and enough local memory to solve the problem. To effectively solve the
$SP_N$ equations for these systems, we take two steps. First, an
algebraic preconditioning strategy is developed for MCSA to address
convergence issues. We utilize a sparse approximate inverse
preconditioner and a sparse matrix-matrix mutliplication kernel to
maintain element-level access to the preconditioned linear operator
and to improve iterative performance. Second, we have parallelized the
Monte Carlo solver by constructing a variant of fully asynchronous
domain decomposition algorithms for Monte Carlo particle transport to
enable solutions on a large number of processors. Using the
preconditioning strategy and domain decomposition algorithm, MCSA
performance is compared with production implementations of Krylov
methods and state-of-the-art algebraic preconditioners for two
eigenvalue problems modeled in three dimensions: a light water reactor
assembly and the C5G7 criticality benchmark. Our studies indicate that
the performance of MCSA when used to solve the $SP_N$ equations for
light water reactor problems in terms of both iterations to converge
and time to solution can be competitive with the preconditioned Krylov
methods when the operator spectrum is properly conditioned using our
sparse approximate inverse strategy. In addition, scalability of the
new domain decomposition algorithm is compared to traditional methods
with the scheme demonstrating good performance on thousands of
processors.

@@

Predictive modeling and simulation capability requires the combination
of high fidelity models, high performance computing hardware that can
handle the intense computational loads required by these models, and
modern algorithms for solving these problems that leverage this high
performance hardware. For nuclear reactor analysis, this predictive
capability can enable tighter design tolerances for improved thermal
performance and efficiency, higher fuel utilization and burnup, and
high confidence in accident scenario models. Modern high fidelity
deterministic methods for large scale problems where the solution over
the entire core of a reactor is desired are often variants on the
discrete ordinates ($S_N$) method \cite{evans_denovo:_2010} or the
method of characteristics \cite{askew_1972}. For fission reactor
neutronics simulations, the discrete ordinates method requires
potentially trillions of unknown angular flux moments to be computed
to achieve good accuracy for the responses of interest
\cite{slaybaugh_acceleration_2011} which in many cases can be
intractable without sufficient computational hardware. On the other
side of the spectrum, low-fidelity diffusion solutions are often
leveraged as an alternative to computing a solution over the entire
domain a reactor core where diffusion theory is often valid.

In the 1960's, Gelbard developed an ad-hoc, multidimensional extension
of the simple one-dimensional planar $P_N$ equations that created a
system of coupled, diffusion-like equations known as the simplified
$P_N$ ($SP_N$) equations \cite{gelbard_1960} which incorporate
additional angular information. Further work in both aysmptotic
analysis and algorithmic development proceeded in the following
decades to serve as a theoretical basis for the method
\cite{morel_1996,brantley_simplified_2000}. Recent work has expanded
the discretization to a fully algebraic form ammenable to solution
with modern linear algebra techniques including Krylov solvers and
advanced preconditioners including multigrid methods and incomplete
factorizations \cite{hamilton_2014}. Using these new techniques, the
reduction in numerical complexity compared to current deterministic
solution methods using the $SP_N$ approximation could mean significant
savings in both compute time and memory required and permit larger
simulations to be performed on the available computing platforms using
standard tools.

Looking towards future computing platforms, as machines begin to
operate at hundreds of petaflops peak performance and beyond, trends
toward reduced energy consumption will require incredibly high levels
of concurrency to achieve the desired computation rates
\cite{kogge_using_2011}. The end result of these hardware changes is
that the large number of low-powered processors populating the machine
will be prone to both soft failures such as bit errors in floating
point operations and hard failures where the data owned by that
processor cannot be recovered
\cite{u.s._department_of_energy_resilient_2012}. Because these
failures are predicted to be common, resilient solver technologies are
a possible means of mitigating the impact of these events at the
application level. With solutions to deterministic transport problems
based on Monte Carlo techniques, such issues are potentially
alleviated by statistical arguments. In the case of soft failures,
isolated floating point errors in the Monte Carlo simulation are
absorbed within tally statistics while a hard failure is treated as a
high variance event where some portion of the Monte Carlo histories
are lost.

The goal of this work is to research and develop Monte Carlo Synthetic
Acceleration (MCSA) methods for neutron transport problems at
scale. These methods are based on using the Monte Carlo method to
directly accelerate the convergence of traditional fixed-point
iterative schemes for these problems and have the potential to improve
the resiliency of large-scale solutions to the neutron transport
problem. We aim to address both the benefits and shortcomings of MCSA
in the context of the physics of interest and desire to identify areas
in which improvements can be made to the algorithm so that the
tradeoff between algorithmic performance and the potential for
resilience at scale is not too great. In previous work
\cite{evans_monte_2014}, we applied MCSA to the thermal radiation
diffusion equation where, for several transient problems, we observed
that MCSA performed competetively with preconditioned Krylov
methods. In a continuation of these studies, we apply the method to
light water reactor problems discretized by the $SP_N$ equations. This
system has no time dependence, a large scattering component, and
assymetry in the linear operator, creating a set of problems
complementary to the transient, opaque, and symmetric systems of our
previous analysis. For problems of this type, our results indicate
that the primitive diagonal preconditioning used in our previous work
is not suitable to achieve convergence with the method. In addition,
the problems studied in this work are large enough that domain
decomposition is required in order to obtain enough memory as well as
time-efficient solutions.

To enable solutions of the $SP_N$ equations for these problems, this
work develops two new techniques. First, we define an algebraic
preconditioning strategy for MCSA that leverages an explicitly formed
preconditioner as well as a matrix-matrix product to construct the
matrix elements of the preconditioned system. As a preconditioner,
this work uses a sparse approximate inverse technique which results in
a sparse matrix with element-wise access, a necessary requirement for
implementation of the Neumann-Ulam Monte Carlo algorithm. Second, we
develop a new domain decomposition scheme for MCSA with a new fully
asynchronous domain decomposed Monte Carlo kernel for the Neumann-Ulam
algorithm. In tandem, these two developments permit solutions of the
$SP_N$ equations for large-scale light water criticality problems. We
believe both of these contributions in preconditioning and
parallelizing the Neumann-Ulam algorithm are new and unique.

The paper is organized as follows. First, we briefly introduce the
$SP_N$ equations for eigenvalue problems along with the discretization
used for this work. We then describe the application of the MCSA
algorithm to the solution of the equations and introduce the
criticality problems used for the numerical studies in the this
work. Next, we demonstrate that the basic Jacobi preconditioning
applied in our previous work is unsatisfactory to achieve convergence
for these problems and subsquently present our new preconditioning
scheme. We then parallelize the MCSA algorithm and describe our new
domain decomposition scheme deisgned to enable very large problems. We
then apply this preconditioning scheme and parallel algorithm to the
criticality problems and compare the performance of the preconditioned
MCSA method to preconditioned Krylov methods. We then finish with
conclusions and indication of future work.

%%---------------------------------------------------------------------------%%
\section{Simplified $P_N$ Approximation}

Steven or Tom....

%%---------------------------------------------------------------------------%%
\section{Monte Carlo Synthetic Acceleration}
\label{sec:mcsa}
In this section we briefly review an adjoint variant the Monte Carlo
method of Neumann and Ulam for linear systems and Monte Carlo
Synthetic Acceleration.

\subsection{Neumann-Ulam Monte Carlo Algorithm}
An alternative approach to approximate matrix inversion is to employ
Monte Carlo methods that sample a distribution with an expectation
value equivalent to that of the inverted operator. Such methods have
been in existence for decades with the earliest reference noted here a
manuscript published in 1950 by Forsythe and Leibler
\cite{forsythe_matrix_1950}. In their outline, Forsythe and Leibler
in fact credit the creation of this technique to J. Von Neumann and
S.M. Ulam some years earlier than its publication. In 1952 Wasow
provided a more formal explanation of Von Neumann and Ulam's method
\cite{wasow_note_1952} and Hammersley and Handscomb's 1964 monograph
\cite{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
\cite{spanier_monte_1969} present additional detail on this topic
using a collection of references from the 1950's and early
1960's which we use here in our discussion.

To derive the algoritm, consider a linear problem:
\begin{equation}
  \ve{A} \ve{x} = \ve{b}\:,
  \label{eq:linear_problem}
\end{equation}
where $\ve{A} \in \mathbb{R}^{N \times N}$, $\ve{x} \in \mathbb{R}^N$,
and $\ve{b} \in \mathbb{R}^N$. We define the linear system adjoint to
Eq~(\ref{eq:linear_problem}):
\begin{equation}
  \ve{A}^T \ve{y} = \ve{d}\:,
  \label{eq:adjoint_linear_problem}
\end{equation}
where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
vectors respectively and $\ve{A}^T$ is the adjoint operator for
$\ve{A} \in \mathbb{R}^{N \times N}$ which we then split to get:
\begin{equation}
  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:,
  \label{eq:adjoint_split_system}
\end{equation}
where $\ve{H}^T = \ve{I} - \ve{A}^T$ is the adjoint \textit{iteration
  matrix}.  We then solve for $\ve{y}$ by writing
Eq~(\ref{eq:adjoint_split_system}) as:
\begin{equation}
  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
  \label{eq:adjoint_split_system_2}
\end{equation}
which in turn yields the Neumann series using the adjoint operators:
\begin{equation}
  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
  \label{eq:adjoint_neumann_series}
\end{equation}

The Monte Carlo algorithm is generated by expanding
Eq~(\ref{eq:adjoint_neumann_series}) into individual terms:
\begin{equation}
  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h^T_{i,i_1} h^T_{i_1,i_2}\ldots h^T_{i_{k-1},i_k}
  d_{i_k}\:,
  \label{eq:adjoint_neumann_solution}
\end{equation}
which can interpreted as a series of transitions between states of the 
adjoint iteration matrix,
\begin{equation}
 \nu = i \rightarrow i_1 \rightarrow \cdots \rightarrow i_{k-1}
 \rightarrow i_{k}\:,
  \label{eq:mc_walk_permutation}
\end{equation}
where $\nu$ is interpreted as a particular sequence instance. If we
consider $\nu$ to be a sequence created by a Markov chain, we can
build probabilities, weights, and a stochastic estimator to play a
Monte Carlo game whos expectation value is the solution to
Eq~(\ref{eq:linear_problem}). To do this, in
Eqs~(\ref{eq:linear_problem}) and (\ref{eq:adjoint_linear_problem}) we
have 3 unknowns: $\ve{x}$, $\ve{y}$ and $\ve{d}$. To eliminate
$\ve{y}$ and $\ve{d}$ and give a system to solve for $\ve{x}$ we
require two constraints. By defining the following adjoint inner
product equivalence \cite{spanier_monte_1969}:
\begin{equation}
  \langle \ve{A}^T \ve{y}, \ve{x} \rangle = \langle \ve{y}, \ve{A}
  \ve{x} \rangle\:.
  \label{eq:adjoint_operator_product}
\end{equation}
the first constraint follows as:
\begin{equation}
  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
  \label{eq:adjoint_vector_relation}
\end{equation}
As a second constraint we select:
\begin{equation}
  \ve{d} = \boldsymbol{\delta}_j\:,
  \label{eq:adjoint_second_constraint}
\end{equation}
where $\boldsymbol{\delta}_j$ is one of a set of vectors in which the
$j^{th}$ component is the Kronecker delta function $\delta_{i,j}$. If
we apply Eq~(\ref{eq:adjoint_second_constraint}) to our first
constraint Eq~(\ref{eq:adjoint_vector_relation}), we get the following
convenient outcome:
\begin{equation}
  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
  \boldsymbol{\delta}_j \rangle = x_j \:,
  \label{eq:inner_product_constraint}
\end{equation}
meaning that if we compute the inner product of the original
right-hand-side of the linear system and the adjoint solution using a
delta function source, we recover one component of the original
solution.

We can build probabilities and weights by thinking in terms of
radiation transport. This adjoint variation of the Neumann-Ulam
algorithm is equivalent to a traditional forward method where the
initial state, $i_0$, of the random walk is determined by sampling
$\ve{b}$ with probabilities:
\begin{equation}
  P_{(i_0=i)}(\nu) = \frac{|b_i|}{||\ve{b}||_1}\:,
  \label{eq:adjoint_source_probability}
\end{equation}
with a random walk starting weight of:
\begin{equation}
  W_0 = ||\ve{b}||_1 \frac{b_i}{|b_i|}\:,
  \label{eq:adjoint_starting_weight}
\end{equation}
which gives the additional useful relation:
\begin{equation}
  b_{i_0} = W_0 P_{(i_0=i)}\:.
  \label{eq:adjoint_source_definition}
\end{equation}
We build weights and probabilities for the random walk using the
adjoint \textit{Neumann-Ulam decomposition} of $\ve{H}$:
\begin{equation}
  \ve{H}^{T} = \ve{P} \circ \ve{W}\:,
  \label{eq:adjoint_neumann_ulam}
\end{equation}
where $P$ is a row-stochastic matrix with each row being a probability
distribution function and $W$ a corresponding weight matrix.  The
probability matrix is defined element-wise as:
\begin{equation}
  p_{ij} = \frac{|h^T_{ij}|}{\sum_j |h^T_{ij}|}\:,
  \label{eq:adjoint_probability}
\end{equation}
such that we expect to select a new state, $j$, from the current state
in the random walk, $i$, by sampling from the distribution function in
row $i$ of the proability matrix. Per
Eq~(\ref{eq:adjoint_neumann_ulam}), the transition weight from state
$i$ to state $j$ is then defined as:
\begin{equation}
  w_{ij} = \frac{h^T_{ij}}{p_{ij}}\:.
  \label{eq:adjoint_weight}
\end{equation}

Next we define an expectation value for the random walks. The
contribution to the solution in state $j$ from a particular random
walk permutation of length $k$ is given by the \textit{collision
  estimator}:
\begin{equation}
  X_{j}(\nu) = \sum_{m=0}^k W_{m} \delta_{i_m,j}\:,
  \label{eq:adjoint_permutation_contribution}
\end{equation}
where the Kronecker delta, defined as the right-hand-side of the
adjoint linear problem in Eq~(\ref{eq:adjoint_second_constraint}),
indicates that the tally contributes only in the current state, $i_m$,
of the random walk. The expectation value using all random walk
permutations is then:
\begin{equation}
  E\{X_j\} = \sum_{\nu} P_{\nu} X_{j}(\nu)\:,
  \label{eq:adjoint_expectation_value}
\end{equation}
with $P_{\nu}$ the probability of a given random walk, $\nu$.  If
Eq~(\ref{eq:adjoint_expectation_value}) is expanded utilizing
Eq~(\ref{eq:adjoint_source_definition}) to insert the source term, we
directly recover a component of the solution to the linear problem:
\begin{equation}
  \begin{split}
    E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N} b_{i_0} p_{i_0,i_1}p_{i_1,i_2}\ldots
    p_{i_{k-1},i_k} w_{i_0,i_1}w_{i_1,i_2}\ldots
    w_{i_{k-1},i_k} \delta_{i_k,j} \\ &= x_{j}\:,
  \end{split}
  \label{eq:adjoint_expectation_expansion}
\end{equation}
therefore, also providing an unbiased Monte Carlo estimate of the
solution. It should be noted here that
Eq~(\ref{eq:adjoint_expectation_expansion}) only computes a single
component of our desired solution vector when really what we desire is
the entire solution vector. In an adjoint Monte Carlo simulation using
this estimator, the $w_{ij}$ elements that are added into the tally
for each state are only selected if/when the random walk currently
resides in that state. Much like a mesh tally in a particle transport
simulation, we have $N$ simultaneous tallies for $\ve{A} \in
\mathbb{R}^{N \times N}$ that will yield the entire solution vector.

Convergence of the Monte Carlo method, previously assumed by many to
be entirely bound by the spectral radius of $\ve{H}^T$
(e.g. \cite{spanier_monte_1969}), was recently identified as a only a
subset of the convergence criteria. In particular, Ji, Mascagni, and
Li demonstrate in \cite{ji_2013} that a necessary and sufficient
condition for convergence of this algorithm is $\rho(\ve{H^*}) < 1$
with $\ve{H^*}$ defined element-wise as:
\begin{equation}
  h^*_{ij} = \frac{(h^T_{ij})^2}{p_{ij}}\:,
  \label{eq:h_star}
\end{equation}
which is indicative that both the structure of $\ve{H}^T$ and the
selection of transition probabilities affect the convergence of the
algorithm.
